
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 1. Define Keywords, JEL
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\PAPERKEYWORDS}{\textbf{Keywords}: Firearm waiting periods; suicide prevention; gun policy; public health; difference-in-differences; event-study design.}
\newcommand{\PAPERJEL}{\textbf{JEL}: I18; I12; K32; J17; H75.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 2. Define Title
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\PAPERTITLE}{\href{}{The Effects of Waiting Periods on Firearm Suicides in the U.S.}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 3. Define Authors contact information
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\AUTHORHADAH}{Hussain Hadah}
\newcommand{\AUTHORHADAHURL}{https://orcid.org/0000-0002-8705-6386}
\newcommand{\AUTHORHADAHINFO}{\href{\AUTHORHADAHURL}{\AUTHORHADAH}: Department of Economics, Tulane University, 6823 St. Charles Ave., New Orleans, LA 70118, United States (e-mail: \href{mailto:hhadah@tulane.edu}{hhadah@tulane.edu}, phone: +1-602-393-8077)}

\newcommand{\AUTHORCOMP}{Gael Compta}
% \newcommand{\AUTHORCOMPURL}{https://orcid.org/0009-0007-8468-7012}
\newcommand{\AUTHORCOMPINFO}{\href{\AUTHORHADAHURL}{\AUTHORCOMP}: Celia Scott Weatherhead School of Public Health and Tropical Medicine, Tulane University, 1440 Canal Street
New Orleans, LA 70112, United States (e-mail: \href{mailto:gael.compta@gmail.com}{gcompta@tulane.edu})}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 4. Define Thanks
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\ACKNOWLEDGMENTS}{
We would like to thank Ali Saffouri for his exceptional work as a research assistant, the participants of the Applied Microeconomics Workshop at the University of Houston, Tulane University, ASHEcon, the Southern Economics Association Conference, the Murphy Institute's health economics and policy seminar, the anonymous referees, and the editor for their helpful feedback. We also thank Patrick Button, Jonathan Holmes, Austin Denteh, Kevin Callison, and Mary Olson for their comments and feedback.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 5. Define Abstract
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\PAPERABSTRACT}{
Suicide is often an impulsive act, and in the United States, nearly one-half of suicides involve a firearm, the most lethal and readily available method. In this paper, using developments in difference-in-differences design, we study the effect of waiting-period laws on firearm suicides and provide the first causal evaluation of such laws. We find that waiting periods reduce firearm suicides for men (-1.3 per 100,000), while the reduction for the overall population (-0.5 per 100,000), adults aged 55 + (-28.5 per 100,000), and white individuals (-23.1 per 100,000) was statistically insignificant. Waiting periods did not cause an increase in non-firearm suicides. The majority of confidence intervals were in the negative range, and the lack of statistically significant estimates does not mean that waiting periods did not cause a decrease in suicides. These findings provide some evidence that even brief delays in firearm access could disrupt the pathway from suicidal ideation to death, suggesting that cooling-off periods may be a crucial policy tool for suicide prevention. \\
\PAPERJEL}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 6. Define citation or availability of latest draft
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \newcommand{\PAPERDOIURL}{https://doi.org/10.1086/711654}
% \newcommand{\PAPERINFO}{ \url{\PAPERDOIURL}.
% }